Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-07-06 manuscript/manuscript.tex
edits

\begin{abstract}
In studies of infant motor development, a crucial research goal is the identification of latent clusters of infants who experience delayed development, as this is a known risk factor for adverse outcomes later in life. However, there are a number of statistical challenges in modeling infant development: the data are typically skewed, exhibit intermittent missingness, and are highly correlated across the repeated measurements collected during infancy. Using data from the Nurture study, a cohort of over 600 mother-infant pairs followed from pregnancy to 12 months postpartum, we develop a flexible Bayesian finite mixture model for the analysis infant motor development. Our model has a number of attractive features. First, we adopt the multivariate skew normal distribution with cluster-specific parameters that accommodate the inherent correlation and skewness in the data. Second, we model the cluster membership probabilities using a novel application of the P\'olya-Gamma data-augmentation scheme, thereby improving predictions of the cluster membership allocations. Lastly, we impute missing responses under the missing at random assumption by drawing from appropriate conditional multivariate skew  normal distributions. Bayesian inference is achieved through straightforward Gibbs sampling, and can be implemented in available software such as \texttt{R}.  Through simulation studies, we show that the proposed model yields improved inferences over models that ignore skewness. In addition, our imputation method yields improvements compared to conventional missing data methods, including multiple imputation and complete or available case analysis. When applied to Nurture data, we identified two distinct development clusters: one characterized by delayed “U-shaped” development and a higher percentage of male infants and another characterized by more steady development and a lower percentage of males. The clusters also differed in terms of key demographic variables, such as infant race and maternal pre-pregnancy body mass index. These findings can aid investigators in targeting interventions during this critical early-life developmental window.
\end{abstract}

@@

The analysis of infant development trends in the Nurture data presents a number of statistical challenges that motivate our proposed model. First, as depicted in Figure 1, the residuals from repeated measures models of Bayley composite scores exhibit skewness even after adjusting for covariates such as race, sex, and birthweight. This suggests that the assumption of conditional normality made by standard repeated measures models is violated, and a distinguishing feature of the data, skewness, is not being accounted for.

@@

A primary goal of the Nurture study is to identify clusters of infants characterized by distinct motor development trajectories throughout the first year of life. To address this aim, we propose a flexible finite mixture model that accommodates relevant features of the data, such as skewness and dependence among the responses. For $i = 1,...,n$, let $\mathbf{y}_{i}=(y_{i1},\ldots,y_{iJ})^T$ be a $J \times 1$ vector of responses (i.e., Baley composite scores) for subject $i$ across the $J$ measurement occasions. For the analysis of the Nurture data, we propose a finite mixture model of the form
\begin{equation}
	\label{eq:mixture}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ki} f(\mathbf{y}_i|\boldsymbol\theta_k),
\end{equation}
where $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,...,K$) and $\pi_{ki}$ is a subject-specific mixing weight representing the probability that subject $i$ belongs to cluster $k$. For now we assume that $K$ is fixed; in Section 4, we discuss model selection strategies for choosing the optimal value of $K$.

To facilitate posterior inference, we introduce a latent cluster indicator variable $z_i$ taking the value $k \in \{1,...,K\}$ with probability $\pi_{ik}$. Conditional on $z_i = k$, we assume $\mathbf{y}_{i}$ is distributed as
\begin{equation}
\mathbf{y}_{i}|(z_i=k) \sim \text{MSN}_J(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k), \label{eq:msndens}
\end{equation}
where $\text{MSN}_J(\cdot)$ denotes the $J$-dimensional multivariate skew normal density, $\boldsymbol\zeta_{ki}$ is a $J \times 1$ vector of subject- and cluster-specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster-specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster-specific scale matrix that captures dependence among the $J$ responses for subject $i$. The vector $\boldsymbol\alpha_k$ has components $\alpha_{kj}$, $j = 1,...,J$, that control the skewness of outcome $j$ in cluster $k$. When $\boldsymbol\alpha_k = \mathbf{0}$, the MSN distribution reduces to the multivariate normal distribution $\text{N}_J(\boldsymbol\zeta_{ki},\boldsymbol\Omega_k)$, where $\boldsymbol\zeta_{ki}$ is a $J \times 1$ mean vector and $\boldsymbol\Omega_k$ is a $J \times J$ covariance matrix (Azzalini and Dalla Valle, 1996).

@@

Of note, the MSN density can be expressed more compactly in terms of the matrix skew normal (MatSN) density (Chen and Gupta 2005). As we will see in Section 3.6, the matrix representation of the MSN distribution admits convenient conjugate prior distributions for the regression parameters and scale matrices, which in turn leads to efficient Gibbs sampling for posterior inference. Let $\mathbf{Y}_k$ be an ${n_k \times J}$ response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n {1}_{(z_i = k)}$ is the number of observations in cluster $k$. From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_k$ is distributed as
\begin{eqnarray}
\mathbf{Y}_{k} &\sim& \text{MatSN}_{n_k \times J}(\mathbf{M}_k,\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)\label{eq:matsn}\\
%\text{vec}(\mathbf{Y}_k) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})^T\\
\text{vec}(\mathbf{M}_k) &=& (\boldsymbol\zeta_{k1}^T,...,\boldsymbol\zeta_{kn_k}^T)^T, \nonumber
\end{eqnarray}
where $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ as in equation (\ref{eq:msnreg}), $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})^T$, $\mathbf{I}_{n_k}$ is the $n_k\times n_k$ identity matrix, and $\boldsymbol\Omega_k$ is the $J\times J$ scale matrix defined above in equation (\ref{eq:msndens}). From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_{k}$, conditional on the $n_k\times 1$ vector of random effects $\mathbf{t}_{k} = (t_1,...,t_{n_k})^T$, is jointly distributed in matrix form as
\begin{equation}
\mathbf{Y}_k | \mathbf{t}_k \sim \text{MatNorm}_{n_k \times J}(\mathbf{M}^*_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),
\end{equation}
where $\text{MatNorm}_{n_k \times J}(\cdot)$ denotes a $n_k\times J$ matrix normal density, $\text{vec}(\mathbf{M}^*_k) = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ is an $n_kJ\times 1$ mean vector, $\mathbf{X}_k$ is an $n_kJ \times Jp$ design matrix, $\boldsymbol\beta_k$ is the $Jp\times 1$ vector of regression coefficients defined in equation (\ref{eq:msnreg}), and $\boldsymbol\Sigma_k$ is the $J\times J$ conditional covariance of $\boldsymbol\epsilon_{ik}$ given in equation (\ref{eq:msnreg}).

\subsection{Multinomial Regression for the Cluster Indicators}
To accommodate heterogeneity in the cluster-membership probabilities, we model $\pi_{ik}$ as a function of covariates using a multinomial logit model 
\begin{eqnarray}
\pi_{ik} = \Pr(z_i = k|\mathbf{w}_i) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{h = 1}^K \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}},~ k=1,\ldots,K,
\end{eqnarray}
where $\mathbf{w}_i$ is an $r\times 1$ vector of subject-level covariates, $\boldsymbol\delta_k$ is a $r\times 1$ vector of regression parameters associated with membership in cluster $k$. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under this model, $z_i|\mathbf{w}_i \sim \text{Multinomial}(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{i1},...,\pi_{1K})$. During MCMC estimation, the cluster labels $z_i$ are updated from their multinomial full conditional distribution and used in the remaining MCMC steps as cluster assignments as detailed later in section (\textbf{what section}). By allowing the cluster probabilities to vary across subjects, our model can be viewed as a \textit{mixture of experts} model in which $\pi_{ki}$ acts as a \textit{gating function} controlling the prior probability of membership in cluster $k$, and $f(\mathbf{y}_i|\boldsymbol\theta_k)$ in Equation \ref{eq:mixture} is the ``expert'' (Bishop 2016).

To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. By using P\'olya--Gamma data augmentation for the multinomial model, we obtain a \textit{P\'olya--Gamma mixture of experts model}, a computationally efficient way to obtain inferences for the mixing weights in the Bayesian setting. A random variable ${w}$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in \mathbb{R}$ if
\begin{equation}
{w} \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)}, \label{eq:pg1}
\end{equation}
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$ . Polson \textit{et al.} (2013) establish that for the $\text{PG}(b,c)$ density, and for $a,\eta \in \mathbb{R}$,

\begin{equation}
\frac{(\text{e}^{\eta})^a}{(1 + \text{e}^{\eta})^b} = 2^{-b}\text{e}^{\kappa \eta} \int_{0}^{\infty} \text{e}^{-{w}\eta^2/2}p({w}|b,c = 0)d{w}. \label{eq:pg2}
\end{equation}
where $\kappa = a - b/2$. Polson \textit{et al.} also show that the conditional distribution $p(w|b,c)$ results from an ``exponential tilting'' of the $\text{PG}(b,0)$ density, thus

@@

To make use of P\'olya--Gamma data augmentation, we first write the full conditional distribution of $\boldsymbol\delta_k$ as the prior for $\boldsymbol\delta_k$ times the multinomial likelihood for all $z_i$. $$p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ki}^{U_{ki}}(1-\pi_{ki})^{1-U_{ki}},$$
where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $\mathbf{U}_k = (U_{k1},...,U_{kn})$, $U_{ik} = {1}_{(z_i = k)}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ki}$ is defined as in Section 3.4. We can rewrite $\pi_{ki}$ as follows
$$\pi_{ki} = P(U_{ki} = 1) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}}{1 + \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}} = \frac{\text{e}^{\eta_{ki}}}{1 + \text{e}^{\eta_{ki}}}$$
where ${c}_{ik} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ and $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}$. We note that the sum $\sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ki} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}} = \log \left ( 1 + \sum_{h \notin \{k,K \}} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}} \right )^T$$. 

@@

	p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) &\propto& p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \left (\frac{\text{e}^{\eta_{ki}}}{1 + \text{e}^{\eta_{ki}}} \right )^{U_{ki}} \left (\frac{1}{1 + \text{e}^{\eta_{ki}}} \right )^{1-U_{ki}} \nonumber\\ &=& p(\boldsymbol\delta_k) \prod_{i = 1}^n \frac{(\text{e}^{\eta_{ki}})^{U_{ki}}}{1 + \text{e}^{\eta_{ki}}} \label{eq:pglogit}

@@

which is essentially a logistic regression likelihood. We can thus apply the P\'olya--Gamma sampler described in Polson et al. (2013) for logistic regression to update each $\boldsymbol\delta_k$ one at a time based on the binary indicators $U_{ki}$. To accomplish this, we first define for $k = 1,...,k$, the $n \times 1$ vector $\mathbf{U}^*_{k} = \left( \frac{U_{k1}-1/2}{w_{k1}},...,\frac{U_{kn}-1/2}{w_{kn}} \right )^T$. Polson et al. (2013) show that, conditional on $\boldsymbol{w} = (w_{k1},...,w_{kn})^T$, $\mathbf{U}^*_{k}$ follows a $\text{N}_n (\boldsymbol\eta_k,\mathbf{O}_k^{-1})$ distribution with mean $\boldsymbol\eta_k = (\eta_{k1},...,\eta_{kn})^T$ and precision matrix $\mathbf{O}_k = \text{Diag}(w_{k1},...,w_{kn})$. Thus, it follows that the full conditional distribution of $\boldsymbol\delta_k$ is given by 

@@

	p(\boldsymbol\delta_k|\mathbf{z},\mathbf{W}, \mathbf{O}_k) \propto p(\boldsymbol\delta_k) \text{exp} \left [ - \frac{1}{2} (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)^T \mathbf{O}_k (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)\right],

@@

where $p(\boldsymbol\delta_k)$ is the prior distribution for $\boldsymbol\delta_k$, which we detail in Section 3.4 As detailed in Section 3.4, if we assume $p(\boldsymbol\delta_k)$ to be multivariate normal, then $\boldsymbol\delta_k$ has a closed-form multivariate normal full conditional distribution that can be easily embedded within our proposed Gibbs sampling routine. 

@@

To accommodate missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness (i.e. observations missing not at random). Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,\mathbf{y}^{obs}_i)$:

@@

\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& \text{N}_{J-q_i}(\boldsymbol\zeta^{cond}_{ki},\boldsymbol\Sigma^{cond}_k),~\text{where}\nonumber\\
\boldsymbol\zeta^{cond}_{ki}&=& \boldsymbol\zeta^{miss}_{ki} + \boldsymbol\Sigma_{k12} \boldsymbol\Sigma_{k22}^{-1}(\mathbf{y}^{obs}_i - \boldsymbol\zeta^{obs}_{ki}) \label{eq:impute}\\
\boldsymbol\Sigma^{cond}_k&=& \boldsymbol\Sigma_{k11} - \boldsymbol\Sigma_{k12}\boldsymbol\Sigma_{k22}^{-1}\boldsymbol\Sigma_{k21},\nonumber \\
\boldsymbol\zeta_{ki} &=& \begin{pmatrix} \boldsymbol\zeta_{ki}^{miss} \\ \boldsymbol\zeta_{ki}^{obs} \end{pmatrix}, \ \text{and} \nonumber \\

@@

Since the primary objective of our proposed model is to provide a flexible framework to account for the inherent features of infant development data such as skewness, correlation among repeated measures, and missingness, we assume the number of clusters $K$ to be specified \textit{a priori}.

>>>>

\begin{abstract}
In studies of infant motor development, a crucial research goal is the identification of latent clusters of infants who experience delayed motor development, as this is a known risk factor for adverse outcomes later in life. However, there are a number of statistical challenges in modeling infant motor development: the data are typically skewed, exhibit intermittent missingness, and are highly correlated across the repeated measurements collected during infancy. Using data from the Nurture study, a cohort of over 600 mother-infant pairs followed from pregnancy to 12 months postpartum, we develop a flexible Bayesian finite mixture model for the analysis infant motor development. Our model has a number of attractive features. First, we adopt the multivariate skew normal distribution with cluster-specific parameters that accommodate the inherent correlation and skewness in the data. Second, we model the cluster membership probabilities using a novel application of the P\'olya-Gamma data-augmentation scheme, thereby improving predictions of the cluster membership allocations. Lastly, we impute missing responses under the missing at random assumption by drawing from appropriate conditional multivariate skew  normal distributions. Bayesian inference is achieved through straightforward Gibbs sampling, and can be implemented in available software such as \texttt{R}.  Through simulation studies, we show that the proposed model yields improved inferences over models that ignore skewness. In addition, our imputation method yields improvements compared to conventional missing data methods, including multiple imputation and complete or available case analysis. When applied to Nurture data, we identified two distinct development clusters: one characterized by delayed “U-shaped” motor development and a higher percentage of male infants and another characterized by more steady motor development and a lower percentage of males. The clusters also differed in terms of key demographic variables, such as infant race and maternal pre-pregnancy body mass index. These findings can aid investigators in targeting interventions during this critical early-life developmental window.
\end{abstract}

@@

The analysis of infant motor development trends in the Nurture data presents a number of statistical challenges that motivate our proposed model. First, as depicted in Figure 1, the residuals from repeated measures models of Bayley composite scores exhibit skewness even after adjusting for covariates such as race, sex, and birthweight. This suggests that the assumption of conditional normality made by standard repeated measures models is violated, and a distinguishing feature of the data, skewness, is not being accounted for.

@@

A primary goal of the Nurture study is to identify clusters of infants characterized by distinct motor development trajectories throughout the first year of life. To address this aim, we propose a flexible finite mixture model that accommodates relevant features of the data, such as skewness and dependence among the responses. For $i = 1,...,n$, let $\mathbf{y}_{i}=(y_{i1},\ldots,y_{iJ})^T$ be a $J \times 1$ vector of responses (i.e., Bayley composite scores) for subject $i$ across the $J$ measurement occasions. For analysis of the Nurture data, we propose a finite mixture model of the form
\begin{equation}
	\label{eq:mixture}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ki} f(\mathbf{y}_i|\boldsymbol\theta_k),
\end{equation}
where $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,...,K$) and $\pi_{ki}$ is a subject-specific mixing weight representing the probability that subject $i$ belongs to cluster $k$. For now we assume that $K$ is fixed; in Section \ref{s:mcmcmisc}, we discuss model selection strategies for choosing the optimal value of $K$.

To facilitate posterior inference, we introduce a latent cluster indicator variable $z_i$ taking the value $k \in \{1,...,K\}$ with probability $\pi_{ki}$. Conditional on $z_i = k$, we assume $\mathbf{y}_{i}$ is distributed as
\begin{equation}
\mathbf{y}_{i}|(z_i=k) \sim \text{MSN}_J(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k), \label{eq:msndens}
\end{equation}
where $\text{MSN}_J(\cdot)$ denotes the $J$-dimensional multivariate skew normal density, $\boldsymbol\zeta_{ki}$ is a $J \times 1$ vector of subject- and cluster-specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster-specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster-specific scale matrix that captures dependence among the $J$ responses for subject $i$. The vector $\boldsymbol\alpha_k$ has components $\alpha_{kj}$, $j = 1,...,J$, that control the skewness of outcome $j$ in cluster $k$. When $\boldsymbol\alpha_k = \mathbf{0}$, the MSN distribution reduces to the multivariate Normal distribution $\text{N}_J(\boldsymbol\zeta_{ki},\boldsymbol\Omega_k)$, where $\boldsymbol\zeta_{ki}$ is a $J \times 1$ mean vector and $\boldsymbol\Omega_k$ is a $J \times J$ covariance matrix (Azzalini and Dalla Valle, 1996).

@@

Of note, the MSN density can be expressed more compactly in terms of the Matrix Skew Normal (MatSN) density (Chen and Gupta 2005). As we will see in Section 3.6, the matrix representation of the MSN distribution admits convenient conjugate prior distributions for the regression parameters and scale matrices, which in turn leads to efficient Gibbs sampling for posterior inference. Let $\mathbf{Y}_k$ be an ${n_k \times J}$ response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n {1}_{(z_i = k)}$ is the number of observations in cluster $k$. From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_k$ is distributed as
\begin{eqnarray}
\mathbf{Y}_{k} &\sim& \text{MatSN}_{n_k \times J}(\mathbf{M}_k,\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)\label{eq:matsn}\\
%\text{vec}(\mathbf{Y}_k) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})^T\\
\text{vec}(\mathbf{M}_k) &=& (\boldsymbol\zeta_{k1}^T,...,\boldsymbol\zeta_{kn_k}^T)^T, \nonumber
\end{eqnarray}
where $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ as in equation (\ref{eq:msnreg}), $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})^T$, $\mathbf{I}_{n_k}$ is the $n_k\times n_k$ identity matrix, and $\boldsymbol\Omega_k$ is the $J\times J$ scale matrix defined above in equation (\ref{eq:msndens}). From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_{k}$, conditional on the $n_k\times 1$ vector of random effects $\mathbf{t}_{k} = (t_1,...,t_{n_k})^T$, is jointly distributed in matrix form as
\begin{equation}
\mathbf{Y}_k | \mathbf{t}_k \sim \text{MatNorm}_{n_k \times J}(\mathbf{M}^*_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),
\end{equation}
where $\text{MatNorm}_{n_k \times J}(\cdot)$ denotes a $n_k\times J$ matrix normal density, $\text{vec}(\mathbf{M}^*_k) = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ is an $n_kJ\times 1$ mean vector, $\mathbf{X}_k$ is an $n_kJ \times Jp$ design matrix, $\boldsymbol\beta_k$ is the $Jp\times 1$ vector of regression coefficients defined in equation (\ref{eq:msnreg}), and $\boldsymbol\Sigma_k$ is the $J\times J$ conditional covariance of $\boldsymbol\epsilon_{ik}$ given in equation (\ref{eq:msnreg}).

\subsection{Multinomial Regression for the Cluster Indicators}
\label{s:multinom}
To accommodate heterogeneity in the cluster-membership probabilities, we model $\pi_{ki}$ as a function of covariates using a multinomial logit model 
\begin{eqnarray}
\label{eq:pik}
\pi_{ki} = \Pr(z_i = k|\mathbf{w}_i) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{h = 1}^K \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}},~ k=1,\ldots,K,
\end{eqnarray}
where $\mathbf{w}_i$ is an $r\times 1$ vector of subject-level covariates, $\boldsymbol\delta_k$ is a $r\times 1$ vector of regression parameters associated with membership in cluster $k$. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under this model, $z_i|\boldsymbol\pi_i \sim \text{Multinomial}(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{1i},...,\pi_{Ki})$. During MCMC estimation, the cluster labels $z_i$ are updated from their multinomial full conditional distribution and used in the remaining MCMC steps as cluster assignments as detailed later in Section \ref{s:postinf}. By allowing the cluster probabilities to vary across subjects, our model can be viewed as a \textit{mixture of experts} model in which $\pi_{ki}$ acts as a \textit{gating function} controlling the prior probability of membership in cluster $k$, and $f(\mathbf{y}_i|\boldsymbol\theta_k)$ in equation (\ref{eq:mixture}) is the ``expert'' providing information on the within-cluster distribution of $\mathbf{y}_i$ (Bishop 2016).

To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson {et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. By using P\'olya--Gamma data augmentation for the multinomial model, we obtain a \textit{P\'olya--Gamma mixture of experts model}, a computationally efficient way to obtain inferences for the mixing weights in the Bayesian setting. A random variable ${w}$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in \mathbb{R}$ if
\begin{equation}
{w} \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)}, \label{eq:pg1}
\end{equation}
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$ . Polson {et al.} establish that for $a,\eta \in \mathbb{R}$,

\begin{equation}
\frac{(\text{e}^{\eta})^a}{(1 + \text{e}^{\eta})^b} = 2^{-b}\text{e}^{\kappa \eta} \int_{0}^{\infty} \text{e}^{-{w}\eta^2/2}p({w}|b,c = 0)d{w}, \label{eq:pg2}
\end{equation}
where $\kappa = a - b/2$ and $p(w|b,c = 0)$ denotes a $\text{PG}(b,0)$ density. Polson {et al.} further show that the conditional distribution $p(w|b,c)$ results from an ``exponential tilting'' of the $\text{PG}(b,0)$ density, thus

@@

Polson et al. use these results to show that, in the logistic model, the Bernoulli likelihood can be written in terms of the kernel of a Normal density and a $\text{PG}(b,0)$ density, resulting in a Normal full conditional distribution for the logistic regression parameters. To extend the P\'olya--Gamma data augmentation approach to the multinomial model, we first introduce the binary indicators $U_{ki}$, such that $U_{ki} = 1_{z_i = k}$ -- i.e. $U_{ki}$ is an indicator variable taking the value $1$ if subject $i$ belongs to cluster $k$, and $0$ otherwise. The full conditional distribution of $\boldsymbol\delta_k$ is then given as  $$p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ki}^{U_{ki}}(1-\pi_{ki})^{1-U_{ki}},$$
where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, and $\pi_{ki}$ is defined as in equation (\ref{eq:pik}). We can rewrite $\pi_{ki}$ in terms of $U_{ki}$ as
$$\pi_{ki} = P(U_{ki} = 1) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}}{1 + \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}} = \frac{\text{e}^{\eta_{ki}}}{1 + \text{e}^{\eta_{ki}}},$$
where ${c}_{ki} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ and $\eta_{ki} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}$. We note that the sum $\sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ki} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}} = \log \left ( 1 + \sum_{h \notin \{k,K \}} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}} \right )^T.$$

@@

	p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) &\propto& p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \left (\frac{\text{e}^{\eta_{ki}}}{1 + \text{e}^{\eta_{ki}}} \right )^{U_{ki}} \left (\frac{1}{1 + \text{e}^{\eta_{ki}}} \right )^{1-U_{ki}} \nonumber\\ &=& p(\boldsymbol\delta_k) \prod_{i = 1}^n \frac{(\text{e}^{\eta_{ki}})^{U_{ki}}}{1 + \text{e}^{\eta_{ki}}}, \label{eq:pglogit}

@@

which is a logistic likelihood of the same form used by Polson et al. to derive the data-augmented P\'olya--Gamma Gibbs sampler for the logistic regression model. We can thus apply the P\'olya--Gamma sampler described in Polson et al. for logistic regression to update each $\boldsymbol\delta_k$ one at a time based on the binary indicators $U_{ki}$. To accomplish this, we first define for $k = 1,...,K$, the $n \times 1$ vector $\mathbf{U}^*_{k} = \left( \frac{U_{k1}-1/2}{w_{k1}},...,\frac{U_{kn}-1/2}{w_{kn}} \right )^T$. Polson et al. show that, conditional on $\boldsymbol{w} = (w_{k1},...,w_{kn})^T$, $\mathbf{U}^*_{k}$ follows a $\text{N}_n (\boldsymbol\eta_k,\mathbf{O}_k^{-1})$ distribution with mean $\boldsymbol\eta_k = (\eta_{k1},...,\eta_{kn})^T$ and precision matrix $\mathbf{O}_k = \text{Diag}(w_{k1},...,w_{kn})$. Thus, it follows that the full conditional distribution of $\boldsymbol\delta_k$ is given by 

@@

	p(\boldsymbol\delta_k|\mathbf{z},\mathbf{W}, \mathbf{O}_k) \propto p(\boldsymbol\delta_k) \text{exp} \left [ - \frac{1}{2} (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)^T \mathbf{O}_k (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)\right].

@@

As detailed in Section 3.4, if we assume $p(\boldsymbol\delta_k)$ to be multivariate normal, then $\boldsymbol\delta_k$ has a closed-form multivariate normal full conditional distribution that can be easily embedded within our proposed Gibbs sampling routine.  For more details on the P\'olya-Gamma Gibbs sampler for logistic and multinomial models, see Polson et al. (2013).

@@

To accommodate intermittent missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' -- that is, as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness (i.e. observations missing not at random). Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate Normal distribution given $(z_i,t_i,\mathbf{y}^{obs}_i)$:

@@

\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& \text{N}_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k),~\text{where}\nonumber\\
\boldsymbol\mu^{cond}_{ki}&=& \boldsymbol\mu^{miss}_{ki} + \boldsymbol\Sigma_{k12} \boldsymbol\Sigma_{k22}^{-1}(\mathbf{y}^{obs}_i - \boldsymbol\mu^{obs}_{ki}) \nonumber \\
\boldsymbol\Sigma^{cond}_k&=& \boldsymbol\Sigma_{k11} - \boldsymbol\Sigma_{k12}\boldsymbol\Sigma_{k22}^{-1}\boldsymbol\Sigma_{k21}, \label{eq:impute} \\
\boldsymbol\mu_{ki} &=& \begin{pmatrix} \boldsymbol\mu_{ki}^{miss} \\ \boldsymbol\mu_{ki}^{obs} \end{pmatrix}, \ \text{and} \nonumber \\

@@

Since the primary objective of our proposed model is to provide a flexible framework to account for the inherent features of infant motor development data such as skewness, correlation among repeated measures, and missingness, we assume the number of clusters $K$ to be specified \textit{a priori}. Often in studies of infant motor development clusters, focus lies on explaining the heterogeneity in motor development trajectories with a parsimonious number of latent clusters. To this end, we propose the use of the ``widely applicable information criterion'' (WAIC) introduced by Watanabe (2010) for model selection. WAIC has the desireable property of penalizing complexity in models -- a feature congruent with our goal of explaining infant motor development heterogeneity with a parsimonious number of clusters. We refer to Gelman et al. (2014) for a detailed discussion of WAIC and comparison to other popular model fit criteria, such as DIC. In practice, we follow the approach taken by Neelon and Chung (2016), wherein a small set of possible values for $K$, e.g. $K \in \{2,3,4 \}$, are specified \textit{a priori} and models fit under each value of $K$ are compared using WAIC. In Section \ref{s:sim} we show with a simulation studies that this approach recovered the true value of $K$ under realistic parameter settings.
